$Z_3$ topological order in the quantum dimer-pentamer model

### 2016-12-03 qdpm_paper.tex
updates to figures
small changes to manuscript

We now consider relaxing the local constraint to a parity constraint such that only $Q_v^{Z_2} \equiv (Q_v \mod 2)$ is required to vanish; the resulting GQDM are the even and odd sectors are the  even \& odd Ising gauge theories (IGT) (and the closely related even and odd sector Toric Codes). In these cases, the local constraint requires $\alpha_v \in \{ 0, \pi\}$ in \Eqref{eq:Gv} and therefore the gauge symmetry is reduced from $U(1)$ to $Z_2$. The set of allowed dimer resonances $\mathcal{C}_{\rm{TC}}$ is expanded accordingly:
\begin{align}
\mathcal{C}_{\rm{TC}} &\equiv \bigl\{ \left(c_{\Box,i},\bar{c}_{\Box,i}\right) \bigr\} \\
&=\bigl\{
\left(\vrtplqt,\hrzplqt\right),\left(\threeleftplqt,\oneleftplqt\right),\left(\twodownplqt,\twoupplqt\right),\left(\bplqt,\fplqt\right) \bigr\}
\end{align}
where we implicitly include rotational symmetry related resonances. We will limit or discussion to $v_\bplqt=0$ such that the corresponding Hamiltonian is
\begin{equation}
\HIGT\left(t\right) \equiv \HGQDM \biggl( \mathcal{C}_{\rm{IGT}},\bigl\{ \left(t,0\right) \bigr\} \biggr).
\end{equation}
In this case the equal superposition RK wavefunction is the exact ground state occurs in the absence of a potential energy term, as we can define a projection operator on each plaquette
\begin{equation}
h_\Box \equiv \frac{1}{2} \left( \mathbb{1} - \sum_i \eket{c_{\Box,i}}\ebra{\bar{c}_{\Box,i}} + h.c. \right)
\end{equation}
such that $\HIGT$ takes the form of \Eqref{eq:HRK} (up to a constant), and the RK state is the ground state for all values of $t$.

@@

We define the Hilbert space of the QDPM to be an odd sector GQDM where vertices are constrained to be touched by a single dimer or the center of a pentamer, such that $n_v = 1,4$. For defect-free configurations in the QDPM, $Q_v$ only vanishes modulo $3$;  thus we may define a $Z_3$ charge,
\begin{equation}
Q_v^{Z_3} \equiv \left(Q_v~\rm{mod}~3\right)
\end{equation}
 which strictly vanishes at each vertex, representing the local constraint. All states in the QDPM are invariant under an exact local $Z_3$ gauge symmetry given by \Eqref{eq:Gv} with $\alpha_v \in \{0,\pm 2\pi/3\}$. Thus introducing pentamers in the QDM reduces the gauge symmetry from $U(1)$ to $Z_3$.

@@

To test for the possibility of translational and rotational symmetry breaking corresponding to the features in $S_d$ at the Brillouin zone edges, we define several order parameters which would capture such potential symmetry breaking. For rotational symmetry breaking we define 
\begin{equation}
\Mrot \equiv  \frac{2}{L^2} \left(N_h - N_v\right) \label{eq:Mrot},
\end{equation}
where $N_v$ ($N_h$) is the total number of vertical (horizontal) dimers; note that $\exval{M_{\rm{rot}}^2 } \neq 0$ implies there is rotational symmetry breaking. to To capture translational symmetry breaking by one lattice site in a direction perpendicular to the majority of the dimers we define $\Mperp$ 
\begin{align}
\Mperp \equiv  \frac{8}{L^2} \sum_{\Box} \Bigl(  &(-1)^{r_x} \theta\left(\Mrot \right) \Pi_\Box^v \notag\\
+ &(-1)^{r_y}\theta\left(-\Mrot \right) \Pi_\Box^h\Bigr) \label{eq:Mperp},
\end{align}
where $\theta(x)$ is the Heaviside function, and $\Pi_\Box^h$ ($\Pi_\Box^v$) is one if the plaquette $\Box$ has two horizontal (vertical) dimers, and zero otherwise. Similarly, we define $\Mpar$ to be given by \Eqref{eq:Mperp} with $r_x$ and $r_y$ swapped between terms. Additionally, we define order parameters that indicate ``columnar" and ``staggered" dimer order~\cite{Syljuasen2005}:
\begin{align}
\MstagSq \equiv  \frac{1}{L^2} \biggl[ &\biggl( \sum_{\ell_h} (-1)^{r_x+r_y} n_d \left(\vec{r}_{\ell_h}\right) \biggr)^2  \notag\\
+&\biggl( \sum_{\ell_v} (-1)^{r_x+r_y} n_d \left(\vec{r}_{\ell_v}\right) \biggr)^2\biggr]\label{eq:Mstag},
\end{align}
\begin{align}
\McolSq \equiv  \frac{1}{L^2} \biggl[  &\biggl( \sum_{\ell_h} (-1)^{r_x} n_d \left(\vec{r}_{\ell_h}\right) \biggr)^2 \notag \\
+ &\biggl( \sum_{\ell_v} (-1)^{r_y} n_d \left(\vec{r}_{\ell_v}\right) \biggr)^2  \biggr]\label{eq:Mcol},
\end{align}
where the sums over $\ell_h$ ($\ell_v$) are taken over all horizontal (vertical) links. In \figref{fig:order_params} we have plotted the finite-size scaling of these order parameters; we find in all cases that while finite size systems display symmetry breaking due to finite-size effects, these order parameters all vanish in the thermodynamic limit, as demonstrated by the extrapolation in \figref{fig:order_params}.
%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[]
    \centering
    \includegraphics[width=1.0\linewidth]{order_params_plot.pdf}
    \caption{The symmetry breaking order parameters.}
    \label{fig:order_params}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

>>>>

We now consider relaxing the local constraint to a parity constraint such that only $Q_v^{Z_2} \equiv (Q_v \mod 2)$ is required to vanish; the resulting GQDM are the even and odd sectors are the  even \& odd Ising gauge theories (IGT) (and the closely related even and odd sector Toric Codes). In these cases, the local constraint requires $\alpha_v \in \{ 0, \pi\}$ in \Eqref{eq:Gv} and therefore the gauge symmetry is reduced from $U(1)$ to $Z_2$. The set of allowed dimer resonances $\mathcal{C}_{\rm{IGT}}$ is expanded accordingly:
\begin{align}
\mathcal{C}_{\rm{IGT}} &\equiv \bigl\{ \left(c_{\Box,i},\bar{c}_{\Box,i}\right) \bigr\} \\
&=\bigl\{
\left(\vrtplqt,\hrzplqt\right),\left(\threeleftplqt,\oneleftplqt\right),\left(\twodownplqt,\twoupplqt\right),\left(\bplqt,\fplqt\right) \bigr\}
\end{align}
where we implicitly include rotational symmetry related resonances. We will limit or discussion to $v_\bplqt=0$ such that the corresponding Hamiltonian is
\begin{equation}
\HIGT\left(t\right) \equiv \HGQDM \biggl( \mathcal{C}_{\rm{IGT}},\bigl\{ \left(t,0\right) \bigr\} \biggr).
\end{equation}
In this case the equal superposition RK wavefunction is the exact ground state occurs in the absence of a potential energy term, as we can define a projection operator on each plaquette
\begin{equation}
h_\Box \equiv \frac{1}{2} \left( \mathbb{1} - \sum_i \eket{c_{\Box,i}}\ebra{\bar{c}_{\Box,i}} + h.c. \right)
\end{equation}
such that $\HIGT$ takes the form of \Eqref{eq:HRK} (up to a constant), and the RK state is the ground state for all values of $t$.

@@

We define the Hilbert space of the QDPM to be a GQDM where vertices are constrained to be touched by a single dimer or the center of a pentamer, such that $n_v = 1,4$. For defect-free configurations in the QDPM, $Q_v$ only vanishes modulo $3$;  thus we may define a $Z_3$ charge,
\begin{equation}
Q_v^{Z_3} \equiv \left(Q_v~\rm{mod}~3\right)
\end{equation}
 which strictly vanishes at each vertex, representing the local constraint. All states in the QDPM are invariant under an exact local $Z_3$ gauge symmetry given by \Eqref{eq:Gv} with $\alpha_v \in \{0,\pm 2\pi/3\}$. Thus introducing pentamers in the QDM reduces the gauge symmetry from $U(1)$ to $Z_3$.

@@

To test for the possibility of translational and rotational symmetry breaking corresponding to the features in $S_d$ at the Brillouin zone edges, we define several order parameters which would capture such potential symmetry breaking. For rotational symmetry breaking we define 
\begin{equation}
\Mrot \equiv  \frac{2}{L^2} \left(N_v - N_h\right) \label{eq:Mrot},
\end{equation}
where $N_v$ ($N_h$) is the total number of vertical (horizontal) dimers; note that $\exval{M_{\rm{rot}}^2 } \neq 0$ implies there is rotational symmetry breaking. to To capture translational symmetry breaking by one lattice site in a direction perpendicular to the majority of the dimers we define $\Mperp$ 
\begin{align}
\Mperp \equiv  \frac{8}{L^2} \sum_{\Box} \Bigl(  &(-1)^{r_x} \theta\left(\Mrot \right) \Pi_\Box^v \notag\\
+ &(-1)^{r_y}\theta\left(-\Mrot \right) \Pi_\Box^h\Bigr) \label{eq:Mperp},
\end{align}
where $\theta(x)$ is the Heaviside function, and $\Pi_\Box^h$ ($\Pi_\Box^v$) is one if the plaquette $\Box$ has two horizontal (vertical) dimers, and zero otherwise. Similarly, we define $\Mpar$ to be given by \Eqref{eq:Mperp} with $r_x$ and $r_y$ swapped between terms. Additionally, we define order parameters that indicate ``columnar" and ``staggered" dimer order~\cite{Syljuasen2005}:
\begin{align}
\MstagSq \equiv  \frac{1}{L^4} \biggl[ &\biggl( \sum_{\ell_h} (-1)^{r_x+r_y} n_d \left(\vec{r}_{\ell_h}\right) \biggr)^2  \notag\\
+&\biggl( \sum_{\ell_v} (-1)^{r_x+r_y} n_d \left(\vec{r}_{\ell_v}\right) \biggr)^2\biggr]\label{eq:Mstag},
\end{align}
\begin{align}
\McolSq \equiv  \frac{1}{L^4} \biggl[  &\biggl( \sum_{\ell_h} (-1)^{r_x} n_d \left(\vec{r}_{\ell_h}\right) \biggr)^2 \notag \\
+ &\biggl( \sum_{\ell_v} (-1)^{r_y} n_d \left(\vec{r}_{\ell_v}\right) \biggr)^2  \biggr]\label{eq:Mcol},
\end{align}
where the sums over $\ell_h$ ($\ell_v$) are taken over all horizontal (vertical) links. In \figref{fig:order_params} we have plotted the finite-size scaling of these order parameters; we find in all cases that while finite size systems display symmetry breaking due to finite-size effects, these order parameters all vanish in the thermodynamic limit, as demonstrated by the extrapolation in \figref{fig:order_params}.
%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[]
    \centering
    \includegraphics[width=1.0\linewidth]{order_params_plot.pdf}
    \caption{The symmetry breaking order parameters.}
    \label{fig:order_params}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%
